---
title: "Corona Virus Testing (US)"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Status of data

```{r timestamp, echo=FALSE, message=FALSE, comment=NA}
writeLines(paste("These graphs were last updated on", format(Sys.time(), "%A %b %d, %Y, %I:%M %p")))
```

## Source of data

The following summarizes COVID-19 statistics obtained from the covidtracking.com API: <https://covidtracking.com/api/>.

```{r data_processing, echo=FALSE, message=FALSE}

library(rjson)
library(ggplot2)
library(RCurl)
library(lubridate)
library(rlist)

#global params
dpival = 150
date_text_size = 6

#define source URLs for data from covidtracking.org API
url_us_daily = "https://api.covidtracking.com/v1/us/daily.json"
eurl_us_daily <- URLencode(url_us_daily)

#load data as JSON
data_us_daily <- fromJSON(getURL(eurl_us_daily))

#get subset of date range to remove the first N days that are very incomplete (these are the last elements in the list of lists)
days_to_remove = 49
data_us_daily = list.subset(data_us_daily, c(1:(length(data_us_daily)-days_to_remove)))

#determine data shape
max_number_of_columns = 0
col_names = names(data_us_daily[[1]])
number_of_rows = length(data_us_daily)

#determine max number of columns
for(i in 1:number_of_rows){
  number_of_columns = length(data_us_daily[[i]])
  if (number_of_columns > max_number_of_columns){
    max_number_of_columns = number_of_columns
  }
}

#if a row doesn't have full columns, pad it out
#for(i in 1:number_of_rows){
#  number_of_columns = length(data_us_daily[[i]])
#  extra_columns_needed = max_number_of_columns - number_of_columns
#  if (extra_columns_needed > 0){
#    for(j in 1:extra_columns_needed){
#      data_us_daily[[i]][number_of_columns+j] = 0
#    }
#  }
#}

#replace nulls with 0
for(i in 1:number_of_rows){
  for(j in 1:max_number_of_columns){
    if (is.null(data_us_daily[[i]][[j]])){
      data_us_daily[[i]][[j]] = 0
    }
  }
}

#convert lists to dataframes
df <- lapply(data_us_daily, function(day) # Loop through each "day"
{
  # Convert each group to a data frame.
  data.frame(matrix(unlist(day), ncol=max_number_of_columns, byrow=T))
})

#now you have a list of data frames, connect them together in one single dataframe
df <- do.call(rbind, df)

#overwrite the old dataframe with the clean one
data_us_daily = df

#add column names
names(data_us_daily) = col_names

#remove factors
for(i in 1:dim(data_us_daily)[2]){
  data_us_daily[,i] = as.matrix(data_us_daily[,i])
}
data_us_daily[,"date"] = as.numeric(data_us_daily[,"date"])
data_us_daily[,"states"] = as.numeric(data_us_daily[,"states"])
data_us_daily[,"positive"] = as.numeric(data_us_daily[,"positive"])
data_us_daily[,"negative"] = as.numeric(data_us_daily[,"negative"])
data_us_daily[,"posNeg"] = as.numeric(data_us_daily[,"posNeg"])
data_us_daily[,"pending"] = as.numeric(data_us_daily[,"pending"])
data_us_daily[,"death"] = as.numeric(data_us_daily[,"death"])
data_us_daily[,"total"] = as.numeric(data_us_daily[,"total"])
data_us_daily[,"totalTestResults"] = as.numeric(data_us_daily[,"totalTestResults"])

#make sure the order has oldest dates first
data_us_daily = data_us_daily[order(data_us_daily[,"date"], decreasing=FALSE),]

#determine missing per day stats (e.g. the number of new tests performed each day)
data_us_daily[,"NewTests"] = 0
current_tests = 0

data_us_daily[,"NewDeaths"] = 0
current_deaths = 0

data_us_daily[,"NewPositives"] = 0
current_positives = 0

data_us_daily[,"NewPosNeg"] = 0
current_posneg = 0

#determine new per day counts from cumulative values
for(i in 1:length(data_us_daily$date)){
  new_tests = data_us_daily[i,"total"] - current_tests
  data_us_daily[i,"NewTests"] = new_tests
  current_tests = data_us_daily[i,"total"]
  
  new_deaths = data_us_daily[i,"death"] - current_deaths
  data_us_daily[i,"NewDeaths"] = new_deaths
  current_deaths = data_us_daily[i,"death"]
  
  new_positives = data_us_daily[i,"positive"] - current_positives
  data_us_daily[i,"NewPositives"] = new_positives
  current_positives = data_us_daily[i,"positive"]
  
  new_posneg = data_us_daily[i,"posNeg"] - current_posneg
  data_us_daily[i,"NewPosNeg"] = new_posneg
  current_posneg = data_us_daily[i,"posNeg"]
}

#determine change in test capacity from previous day
data_us_daily[,"TestCapacityChange"] = 0
current_capacity = 0
for(i in 1:length(data_us_daily$date)){
  test_capacity_change = data_us_daily[i,"NewTests"] - current_capacity
  data_us_daily[i,"TestCapacityChange"] = test_capacity_change
  current_capacity = data_us_daily[i,"NewTests"]
}

#determine daily and cumulative rates
data_us_daily[,"DailyPosTestRate"] = data_us_daily[,"NewPositives"] / data_us_daily[,"NewPosNeg"]
data_us_daily[,"CumPosTestRate"] = data_us_daily[,"positive"] / data_us_daily[,"posNeg"]
data_us_daily[,"CumDeathRate"] = data_us_daily[,"death"] / data_us_daily[,"positive"]

#create a dataframe of daily death rates
min_date = min(data_us_daily$date)
max_date = max(data_us_daily$date)


#death stats from 2017 CDC
heart_df = data.frame( x = c(min_date, max_date), y = 647457, cause = "Heart disease")
cancer_df = data.frame( x = c(min_date, max_date), y = 599108, cause = "Cancer")
accidents_df = data.frame( x = c(min_date, max_date), y = 169936, cause = "Accidents")
respiratory_df = data.frame( x = c(min_date, max_date), y = 160201, cause = "Respiratory disease")
stroke_df = data.frame( x = c(min_date, max_date), y = 146383, cause = "Stroke")
alzheimers_df = data.frame( x = c(min_date, max_date), y = 121404, cause = "Alzheimer's disease")
diabetes_df = data.frame( x = c(min_date, max_date), y = 83564, cause = "Diabetes")
flu_df = data.frame( x = c(min_date, max_date), y = 55672, cause = "Influenza and pneumonia")
kidney_df = data.frame( x = c(min_date, max_date), y = 50633, cause = "Kidney diseases")
suicide_df = data.frame( x = c(min_date, max_date), y = 47173, cause = "Suicide")
liver_df = data.frame( x = c(min_date, max_date), y = (30223+3290+1047+725+6286), cause = "Liver diseases")
septicemia_df = data.frame( x = c(min_date, max_date), y = (30198+6568+295+879+2869), cause = "Septicemia")
hypertension_df = data.frame( x = c(min_date, max_date), y = (24465+6273+190+1472+2799), cause = "Hypertension")
assault_df = data.frame( x = c(min_date, max_date), y = (5747+9908+266+347+3186), cause = "Assault")

other_causes_of_death = rbind(heart_df, cancer_df, accidents_df, respiratory_df, stroke_df, alzheimers_df, diabetes_df, flu_df, kidney_df, suicide_df, liver_df, septicemia_df, hypertension_df, assault_df)

#adjust for change in population since 2017, based on US population estimate from: https://www.worldometers.info/world-population/us-population/
population_increase_adjustment = 331002651/325084756
other_causes_of_death$y = other_causes_of_death$y*population_increase_adjustment

max_annual_deaths = max(data_us_daily$NewDeaths, (other_causes_of_death$y))
max_daily_deaths = max(data_us_daily$NewDeaths, (other_causes_of_death$y)/365)

ordered_names=c("Heart disease","Cancer","Accidents","Respiratory disease","Stroke","Alzheimer's disease","Diabetes","Influenza and pneumonia","Kidney diseases","Suicide","Liver diseases","Septicemia","Hypertension","Assault")

other_causes_of_death$cause = factor(other_causes_of_death$cause, levels=ordered_names)

```

## Visualizations

### Tests performed

<br>
**Figure 1**. To assess the increasing COVID-19 testing in the US, the number of **new** tests reported each day are summarized below. 

```{r tests_per_day, echo=FALSE, dpi=dpival}

#plot the number of new tests performed each day
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=NewTests, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Number of new COVID-19 tests performed per day in US")
p = p + ylab("New tests performed") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>
**Figure 2**. The cumulative number of tests performed to date. 

```{r cum_tests, echo=FALSE, dpi=dpival}

#plot the cumulative number of tests performed at the end of each day
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=total, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Total cumulative COVID-19 tests performed to date")
p = p + ylab("Cumulative tests performed") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d")
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

**Figure 3**. Change in test capacity from previous day

```{r test_capacity, echo=FALSE, dpi=dpival}

#plot the change in test capacity from the previous day
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=TestCapacityChange, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Change in COVID-19 test capacity from day to day")
p = p + ylab("Change in tests performed compared to previous day") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d")
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

### Positive cases

**Figure 4**. The number of **new** positive cases reported each day are summarized below.

```{r pos_cases, echo=FALSE, dpi=dpival}
#plot the number of positive cases each day
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=NewPositives, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("New positive COVID-19 cases observed per day in US")
p = p + ylab("Daily positive count") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d")
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

**Figure 5**. The cumulative number total positive cases reported to date.

```{r cum_pos_cases, echo=FALSE, dpi=dpival}
#plot the cumulative number of positive cases to date
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=positive, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Total cumulative COVID-19 positive cases observed to date")
p = p + ylab("Cumulative positive count") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

### Positive test rate

**Figure 6**. This graph shows the rate of positive tests observed out of all the tests performed on that same day. As the testing capacity increases we might start to see more people getting test who have less obvious symtoms, more people being tested out of caution, or to assure safety for working with the public, etc.  If that is the case we might see the positive test rate start to drop as testing become widely available.  On the other hand, if infection in the community is already widespread we might not see as much of a drop here.

```{r pos_test_rate, echo=FALSE, dpi=dpival}
#plot the positive test rate each day (new positive / new positive+negative)
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=DailyPosTestRate*100, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Positive COVID-19 test rate observed per day in US")
p = p + ylab("Daily positive test rate % (new pos / new pos+neg)*100") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
print(p)
```

******
<br>

**Figure 7**. This graph is similar to the previous one but shows the cumulative positive test rate. We should see a similar trend here that as the pandemic passes and testing becomes widely available, the number of negative results will accumulate while the number of positive results slows. Over time, if the outbreak is contained this number should drop.  On the other hand, if the outbreak continues to spread, many positives will continue to accumulate. If the rate of community infection is already high or continues to grow, this rate could actually still increase in the coming weeks. 

```{r cum_pos_test_rate, echo=FALSE, dpi=dpival}
#plot the cumulative positive test rate for each day (positive / positive+negative)
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=CumPosTestRate*100, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Cumulative COVID-19 positive test rate observed in US")
p = p + ylab("Cumulative positive test rate % (pos / pos+neg)*100") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
print(p)
```

******
<br>

### Deaths

**Figure 8**. Number of new deaths observed each day.  Pretty self explanatory. We want this to trend downwards...

```{r new_deaths, echo=FALSE, dpi=dpival}
#plot the new deaths each day 
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=NewDeaths, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("New COVID-19 deaths observed in US")
p = p + ylab("New deaths") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

**Figure 9**. Number of new deaths observed each day compared to estimated number of deaths per day from other common causes. Other causes of death are based on the CDC leading causes of death report for 2017 (annual death number / 365 to determine the average daily number for each cause). Also note that for some of the lower causes of death the CDC report does not provide a total count and I compiled it from the ethnicity comparison (table D). This will underestimate deaths from liver disease, septicemia, hypertension, and assault. The estimated deaths from other causes have been adjusted to account for ~2% population increase since 2017.

```{r new_deaths_ref, echo=FALSE, dpi=dpival}
#plot the new deaths each day with reference points from other causes of death 
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=NewDeaths, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("New COVID-19 deaths observed in US compared to other causes")
p = p + ylab("New deaths") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'), limits=c(0,max_daily_deaths))
p = p + geom_hline(data=other_causes_of_death, aes(group=cause, yintercept=y/365, color=cause))
p = p + labs(color = "Cause of Death")
print(p)
```

******
<br>

**Figure 10**. Cumulative deaths observed to date.

```{r deaths, echo=FALSE, dpi=dpival}
#plot the deaths each day 
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=death, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Total cumulative COVID-19 deaths observed in US")
p = p + ylab("Total cumulative deaths") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'))
print(p)
```

******
<br>

**Figure 11**. Cumulative deaths observed to date compared to other causes of death (annual numbers). This has similar caveats to the plot above comparing daily deaths from covid to other common causes.

```{r deaths_ref, echo=FALSE, dpi=dpival}
#plot the deaths each day 
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=death, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Total cumulative COVID-19 deaths compared to other causes")
p = p + ylab("Total cumulative deaths") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
p = p + scale_y_continuous(labels = scales::comma_format(big.mark = ',', decimal.mark = '.'), limits=c(0,max_annual_deaths))
p = p + geom_hline(data=other_causes_of_death, aes(group=cause, yintercept=y, color=cause))
p = p + labs(color = "Cause of Death")
print(p)
```

******
<br>

### Death rate

**Figure 12**. The overall cumulative apparent death rate. An overall estimate of the deadliness of the disease. As testing becomes more widespread and larger numbers of people are tested (even those with less severe symptoms) this number will hopefully drop. In the early days of the outbreak, many of those tested were presumably those with the worst symptoms. This contributes to a very high apparent fatality rate.  As testing becomes more widespread, this rate should drop.  However, unless testing becomes very widespread and systematic, it will probably always overestimate the true lethality of the disease because some will get the virus, remain healthy and never get tested.

```{r cum_death_rate, echo=FALSE, dpi=dpival}
#plot the cumulative death rate (cumulative deaths / cumulative positives)
p = ggplot(data=data_us_daily, aes(x=ymd(date), y=CumDeathRate*100, group=1))
p = p + geom_line()
p = p + geom_point()
p = p + theme_bw()
p = p + theme(axis.text.x=element_text(angle=45, hjust=1, size=date_text_size)) + theme(plot.title = element_text(hjust = 0.5))
p = p + ggtitle("Cumulative COVID-19 death rate observed in US")
p = p + ylab("Observed death rate % (total deaths / total pos)*100") + xlab("Date")
p = p + scale_x_date(date_breaks = "2 day", date_labels =  "%a %b %d") 
print(p)
```

## Executive summary

The following pulls some reference values from the data based on the latest update a summarizes a few key numbers:

```{r executive_summary, echo=FALSE, message=FALSE, comment=NA}
last_day = dim(data_us_daily)[1]

cum_death_rate = round(data_us_daily[last_day,"CumDeathRate"]*100, digits=1)
new_tests = prettyNum(data_us_daily[last_day,"NewTests"], big.mark=",",scientific=FALSE)
total_tests = prettyNum(data_us_daily[last_day,"total"], big.mark=",",scientific=FALSE)
positive_rate = round(data_us_daily[last_day,"DailyPosTestRate"]*100, digits=1)
new_positives = prettyNum(data_us_daily[last_day,"NewPositives"], big.mark=",",scientific=FALSE)
total_cases = prettyNum(data_us_daily[last_day,"positive"], big.mark=",",scientific=FALSE)
cum_pos_test_rate = round(data_us_daily[last_day,"CumPosTestRate"]*100, digits=1)
new_deaths = prettyNum(data_us_daily[last_day,"NewDeaths"], big.mark=",",scientific=FALSE)
deaths = prettyNum(data_us_daily[last_day,"death"], big.mark=",",scientific=FALSE)

x = round(mean(data_us_daily[(dim(data_us_daily)[1]-7):(dim(data_us_daily)[1]),"NewDeaths"]), digits=0)
avg_daily_death_last_7days = prettyNum(x, big.mark=",",scientific=FALSE)
annual_deaths_estimate_last_7days = prettyNum(x*365, big.mark=",",scientific=FALSE)

x = round(mean(data_us_daily[(dim(data_us_daily)[1]-14):(dim(data_us_daily)[1]),"NewDeaths"]), digits=0)
avg_daily_death_last_14days = prettyNum(x, big.mark=",",scientific=FALSE)
annual_deaths_estimate_last_14days = prettyNum(x*365, big.mark=",",scientific=FALSE)

x = round(mean(data_us_daily[(dim(data_us_daily)[1]-21):(dim(data_us_daily)[1]),"NewDeaths"]), digits=0)
avg_daily_death_last_21days = prettyNum(x, big.mark=",",scientific=FALSE)
annual_deaths_estimate_last_21days = prettyNum(x*365, big.mark=",",scientific=FALSE)

x = round(mean(data_us_daily[(dim(data_us_daily)[1]-28):(dim(data_us_daily)[1]),"NewDeaths"]), digits=0)
avg_daily_death_last_28days = prettyNum(x, big.mark=",",scientific=FALSE)
annual_deaths_estimate_last_28days = prettyNum(x*365, big.mark=",",scientific=FALSE)

death_ranking = length(which(unique(other_causes_of_death$y) - max(data_us_daily$death) > 0)) + 1

summary_line1 = paste(new_tests, " tests were performed yesterday, of which ", positive_rate, "% were positive (", new_positives," new cases). Total number of tests performed to date stands at ", total_tests, ". Total cases in the US now stands at ", total_cases, " and the cumulative positive test rate stands at ", cum_pos_test_rate, "%. There were ", new_deaths, " new deaths yesterday, the total number of deaths in the US now stands at ", deaths, " and the cumulative death rate now stands at ", cum_death_rate, "%.", sep="")
writeLines(summary_line1)


summary_line2 = paste("Currently COVID19 is estimated to be the number ", death_ranking, " cause of death when comparing current deaths to expected annual deaths from other causes."  , sep="")
writeLines(summary_line2)

summary_line3 = paste("Over the past week the average daily deaths was ", avg_daily_death_last_7days, ". That rate would result in ", annual_deaths_estimate_last_7days, " deaths annually.", sep="")
writeLines(summary_line3)

summary_line4 = paste("Over the past two weeks the average daily deaths was ", avg_daily_death_last_14days, ". That rate would result in ", annual_deaths_estimate_last_14days, " deaths annually.", sep="")
writeLines(summary_line4)

summary_line5 = paste("Over the past three weeks the average daily deaths was ", avg_daily_death_last_21days, ". That rate would result in ", annual_deaths_estimate_last_21days, " deaths annually.", sep="")
writeLines(summary_line5)

summary_line6 = paste("Over the past four weeks the average daily deaths was ", avg_daily_death_last_28days, ". That rate would result in ", annual_deaths_estimate_last_28days, " deaths annually.", sep="")
writeLines(summary_line6)


```


